---
title: typetracer
output: md_document
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/mpadge/typetracer/workflows/R-CMD-check/badge.svg)](https://github.com/mpadge/typetracer/actions)
[![codecov](https://codecov.io/gh/mpadge/typetracer/branch/main/graph/badge.svg)](https://codecov.io/gh/mpadge/typetracer)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# typetracer

Trace function parameter types in R packages.

```{r}
library (typetracer)
```


## Example #1

Define a function, and use `inject_tracer` to inject parameter tracers used to
trace parameter types on each call. The following function includes an
additional parameter, `z`, which is left undefined here, along with `...` to
allow passing of arbitrary parameter values.

```{r inject}
f <- function (x, y, z, ...) {
    x * x + y * y
}
inject_tracer (f)
```

Calls to the function, `f`, will then trace each parameter of the function. The
current demonstration-only version extracts values for `storage.type` and
`length`. These values can be accessed with the `load_traces` function, with
this example additionally demonstrating:

1. That the additional parameter, `z`, is not traced when left undefined; and
2. That additional parameters passed via `...` are successfully traced.

```{r trace1}
val <- f (x = 1:2, y = 3:4 + 0., a = "blah", b = list (a = 1, b = "b"))
load_traces ()
```

Traces themselves are saved in the temporary directory of the current R
session, and the `load_traces()` function simple loads all traces created in
that session. The function `clear_traces()` removes all traces, so that
`load_traces()` will only load new traces produced after that time.

## Example #2

This section presents a more complex example tracing parameters for a selection
of functions from the base R package.

### Select functions and extract example code

Start with a selection of functions:

```{r base-fns}
pkg <- "base"
fns <- ls (paste0 ("package:", pkg), all.names = TRUE)
pkg_env <- as.environment (paste0 ("package:", pkg))
is_prim <- vapply (fns, function (i) is.primitive (get (i, envir = pkg_env)),
                   logical (1L))
fns <- fns [which (!is_prim)]
# Then reduce to a small sample of those:
fns <- grep ("^[[:alpha:]]", fns, value = TRUE) [1:30]
```

Then extract the example code for those functions from the `.Rd` database of
the base package. For extra simplicity, the functions are reduced to only those
with matching `.Rd` names.

```{r base-rd-db}
rd <- tools::Rd_db (pkg)
fns <- fns [which (fns %in% gsub ("\\.Rd$", "", names (rd)))]
rd <- rd [which (gsub ("\\.Rd$", "", names (rd)) %in% fns)]
```

The following lines extract the example code from those `.Rd` entries:

```{r base-rd-ex}
exs <- lapply (rd, function (i) {
    f <- tempfile ()
    tools::Rd2ex (i, out = f)
    if (!file.exists (f)) { # no examples
        return (NULL)
    }
    
    out <- brio::read_lines (f)
    file.remove (f)
    return (out)
})
exs <- unname (do.call (c, exs))
```

### Inject type tracers in all of those functions

The `system.time` call at the start of the following code demonstrations that
injection is almost instantaneous.

```{r base-inject}
system.time ({
    for (f in fns) {
        f <- get (f, envir = pkg_env)
        inject_tracer (f)
    }
})
```

### Run the example code

Examples can be run through a simple `eval` call, but this must generally be
wrapped like in the following example to suppress generation of plots and
warning messages.

```{r base-eval}
pdf (file = NULL)
o <- suppressWarnings (
    res <- eval (parse (text = exs))
    )
chk <- dev.off ()
```

### Load the traces

Finally, load the resultant type traces as above

```{r base-traces}
load_traces ()
```
