---
title: typetracer
output: md_document
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/mpadge/typetracer/workflows/R-CMD-check/badge.svg)](https://github.com/mpadge/typetracer/actions)
[![codecov](https://codecov.io/gh/mpadge/typetracer/branch/main/graph/badge.svg)](https://codecov.io/gh/mpadge/typetracer)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# typetracer

`typetracer` is an R package to trace function parameter types. The main usage
of `typetracer` is to identify parameters used as input to R functions. Many
computer languages have formal type systems, meaning the types of parameters
must be formally declared and encoded. R is different, and offers no way to
specify the expected types of input parameters. `typetracer` identifies the
types of parameters passed to R functions. The package can trace individual
functions or entire packages, as demonstrated below.


## Installation

The package can be installed with the following command:

```{r remotes, eval = FALSE}
remotes::install_github ("mpadge/typetracer")
```

Then loaded for use by calling `library`:

```{r}
library (typetracer)
```


## Example #1 - A Single Function

`typetracer` works by "injecting" tracing code into the body of a function
using [the `inject_tracer()`
function](https://mpadge.github.io/typetracer/reference/inject_tracer.html).
The following function includes four parameters, including `...` to allow
passing of additional and entirely arbitrary parameter types and values.

```{r inject}
f <- function (x, y, z, ...) {
    x * x + y * y
}
inject_tracer (f)
```

After injecting the `typetracer` code, calls to the function, `f`, will "trace"
each parameter of the function, by capturing both unevaluated and evaluated
representation at the point at which the function is first called. These values
can be accessed with [the `load_traces`
function](https://mpadge.github.io/typetracer/reference/load_traces.html),
which returns a `data.frame` object (in [`tibble`
format](https://tibble.tidyverse.org) with one row for each parameter from each
function call.

```{r trace1}
val <- f (x = 1:2, y = 3:4 + 0., a = "blah", b = list (a = 1, b = "b"))
x <- load_traces ()
x
```

That results shows that all parameters of the function, `f()`, were
successfully traced, including the additional parameters, `a` and `b`, passed
as part of the `...` argument. Such additional parameters can be identified
through having a `"formal"` entry of `NULL`, indicating that they are not part
of the formal arguments to the function.

```{r}
x$uneval [x$par_name %in% c ("a", "b")]
x$eval [x$par_name %in% c ("a", "b")]
```

Traces themselves are saved in the temporary directory of the current R
session, and [the `load_traces()`
function](https://mpadge.github.io/typetracer/reference/load_traces.html)
simply loads all traces created in that session. [The function
`clear_traces()`](https://mpadge.github.io/typetracer/reference/clear_traces.html)
removes all traces, so that
[`load_traces()`](https://mpadge.github.io/typetracer/reference/load_traces.html)
will only load new traces produced after that time.

### Uninjecting Traces

It is important after applying [the `inject_tracer()`
function](https://mpadge.github.io/typetracer/reference/inject_tracer.html) to
restore the functions back to their original form through calling [the obverse
`uninject_tracer()`
function](https://mpadge.github.io/typetracer/reference/uninject_tracer.html).
For the function, `r`, above, this simply requires,

```{r}
uninject_tracer (f)
```


## Example #2 - Tracing a Package

This section presents a more complex example tracing all function calls from
[the `rematch` package](https://github.com/MangoTheCat/rematch), chosen because
it has less code than almost any other package on CRAN. The following single
line traces function calls in all examples for the nominated package. [The
`trace_package()`
function](https://mpadge.github.io/typetracer/reference/trace_package.html)
automatically injects tracing code into every
function within the package, so there is no need to explicitly call [the
`inject_tracer()`
function](https://mpadge.github.io/typetracer/reference/inject_tracer).

```{r trace-rematch, message = FALSE}
res <- trace_package ("rematch")
res
```

The result contains one line for every parameter passed to every function call
in the examples. [The `trace_package()`
function](https://mpadge.github.io/typetracer/reference/trace_package.html)
also includes an additional parameter, `types`, which defaults to `c
("examples", "tests")`, so that traces are also by default generated for all
tests included with local source packages.

The final two columns of the result hold the unevaluated and evaluated
representations of each parameter. The first two values of each demonstrate the
difference:

```{r rematch-output}
res$uneval [1:2]
res$eval [1:2]
```

The example first assigns a variable `isodaten` to the first of the evaluated
values, and then calls the function with `pattern = isodaten`. The second
constructs the vector called `dates` with the second of the evaluated values,
then calls the function with `test = dates`.

### Example #2(a) - Specifying Functions to Trace

[The `trace_package()`
function](https://mpadge.github.io/typetracer/reference/trace_package.html)
also accepts an argument, `functions`, specifying which functions from a
package should be traced. For example,

```{r trace-stats, eval = FALSE}
x <- trace_package ("stats", functions = c ("sd", "var"))
```
```{r stats-output, echo = FALSE}
par_name <- c ("x", "na.rm", "x", "y", "na.rm",
               "use", "x", "y", "na.rm", "use")
storage.mode <- c ("integer", "logical", "integer", "NULL", "logical",
                   "NULL", "integer", "NULL", "logical", "NULL")
# Create an empty list for formal params. "empty" means an empty name or symbol
# object, which can be conveniently constructed with 'substitute()':
formal <- pairlist (
    x = substitute (),
    na.rm = FALSE,
    y = NULL,
    use = substitute ()
)
formal <- formal [match (par_name, names (formal))]

uneval <- as.list (letters [1:10])
names (uneval) <- par_name
eval <- pairlist (
    x = 1:2,
    na.rm = FALSE,
    x = 1:2,
    y = NULL,
    na.rm = FALSE,
    use = NULL,
    x = 1:10,
    y = NULL,
    na.rm = FALSE,
    y = NULL
)


x <- tibble::tibble (
    fn_name = c (rep ("sd", 2L), rep ("var", 8L)),
    fn_call_hash = c (
        rep ("nNbBzFCg", 2L), rep ("kwTgNeUZ", 4L), rep ("risTCGmw", 4L)),
    par_name = c ("x", "na.rm", "x", "y", "na.rm",
                  "use", "x", "y", "na.rm", "use"),
    class = I (as.list (storage.mode)),
    storage.mode = storage.mode,
    length = as.integer (c (2, 1, 2, 0, 1, 0, 10, 0, 1, 0)),
    formal = formal,
    uneval = I (uneval),
    eval = I (as.list (eval))
)
x
```



## Examples #3 - Non-standard Evaluation

This example briefly illustrates some examples of tracing parameters evaluated
in non-standard ways. This first examples demonstrates that parameter values
are captured at the initial point of function entry.

```{r nse1}
eval_x_late_NSE <- function (x, y) {
    y <- 10 * y
    eval (substitute (x))
}
inject_tracer (eval_x_late_NSE)
eval_x_late_NSE (y + 1, 2:3)
res <- load_traces ()
res$par_name
res$uneval
res$eval
```

The parameter `x` is evaluated at the point of function entry as `y + 1` which,
with a value of `y = 2:3`, gives the expected evaluated result of `x = 3:4`,
while the function ultimately returns the expected values of `(10 * 2:3) + 1`,
or `21 31`, because the first line of `y <- 10 * y` is evaluated prior to
substituting the value passed for `x` of `y + 1`.

The second example specifies a default value of `x = y + 1`, with the actual
call passing no value, and thus having `"NULL"` in the unevaluated version,
while evaluated versions remain identical.

```{r nse2}
clear_traces () # clear all preceding traces
eval_x_late_standard <- function (x = y + 1, y, z = y ~ x) {
    y <- 10 * y
    x
}
inject_tracer (eval_x_late_standard)
eval_x_late_standard (, 2:3)
res <- load_traces ()
res$par_name
res$uneval
res$eval
```

The traces produced by `typetracer` also include a column, `formal`, which
contains the default values specified in the definition of
`eval_x_late_standard()`:

```{r}
res$formal
```

Those three columns of `formal`, `uneval`, and `eval` thus contain
all definitions for all parameters passed to the function environment, in the three possible states of:

1. Formal or default values (by definition, in an unevaluated state);
2. The unevaluated state of any specified parameters; and
3. The equivalent versions evaluated within the function environmental.
